---
title: "07-motifs"
format: html
editor: visual
---

Rpubs link:

# Loading R package seqinr and getting an output fasta file

First we need to install the R package seqinr.

```{r}
library(seqinr)

setwd("/home/shared/16TB_HDD_01/fish546/renee/build/megahit_F2R_KM41_out")

# Replace 'input.fasta' with the name of the multi-sequence fasta file
input_file <- "final.contigs.fa"
sequences <- read.fasta(input_file)
```

```{r}
# Set the seed for reproducibility (optional)
set.seed(42)

number_of_sequences_to_select <- 10

if (length(sequences) < number_of_sequences_to_select) {
  warning("There are fewer than 10 sequences in the fasta file. All sequences will be selected.")
  number_of_sequences_to_select <- length(sequences)
}

selected_indices <- sample(length(sequences), number_of_sequences_to_select)
selected_sequences <- sequences[selected_indices]
```

```{r}
# Replace 'output.fasta' with the desired output file name
output_file <- "output/10-seqs.fasta"
write.fasta(selected_sequences, names(selected_sequences), output_file, open = "w")
```

# Creating an index for the 10-sequence fasta

## Creating a necessary index for IGV viewing

```{bash}
#likely will not need; fix issue where gff and fa name did not match
sed -i 's/>lcl|/>/g' output/10-seqs.fasta
```

Let's check things to see that we're on the right track.

```{r engine='bash', eval=TRUE}
head -1 output/10-seqs.fasta.fai
```

We get an output that looks like: k141_28203 601 12 60 61

This means: Sequence name: k141_28203 Sequence length: 601 bases Sequence data starts at byte 16 in the FASTA file Each line contains 60 bases Each line is 61 bytes long (h/t to Lara's .qmd code file)

```{bash}
#needed downstream for IGV
/home/shared/samtools-1.12/samtools faidx \
output/10-seqs.fasta
```

## Finding the motifs

```{bash}
fuzznuc -sequence output/10-seqs.fasta -pattern CG -rformat gff -outfile CGoutput.gff
```
And let's check to see that everything looks good. 
```{r engine='bash', eval=TRUE}
head -6 CGoutput.gff
```
We get something like: 

```
Show in New Window
Search for patterns in nucleotide sequences
Show in New Window
##gff-version 3
##sequence-region k141_28203 1 601
#!Date 2025-05-15
#!Type DNA
#!Source-version EMBOSS 6.6.0.0
k141_28203	fuzznuc	nucleotide_motif	3	4	2	+	.	ID=k141_28203.1;note=*pat pattern:CG
```





